---
title: "Quantification of the association of bacterial clones with antibiotic resistance"
subtitle: "Supplementary Document 1 for 'Rapid heuristic inference of antibiotic resistance and susceptibility by genomic neighbor typing'"
author: "Karel Břinda, Alanna Callendrello, Kevin C Ma, Derek R MacFadden, Themoula Charalampous, Robyn S Lee, Lauren Cowley, Crista B Wadsworth, Yonatan H Grad, Gregory Kucherov, Justin O’Grady, Michael Baym, and William P Hanage"
header-includes:
   - \usepackage{bbm}
output:
  pdf_document:
    fig_caption: yes
    number_sections: TRUE
    keep_tex: TRUE
bibliography: "references.bib"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
---

```{r, include=FALSE}
# todo: add numbers to pg markers

library(knitr)
knitr::opts_chunk$set(
               fig.width=4.5,
               fig.height=4.5,
               fig.align='center',
               fig.pos='t!')

knitr::opts_chunk$set(
               echo=FALSE,
               warning=FALSE,
               message=FALSE,
               cache=TRUE,
               results=FALSE)
```

```{r}
library("knitcitations")
cleanbib()
options("citation_format" = "pandoc")
```

```{r}
source("evaluate-association.R")

library(png)
library(grid)

df.pneumo <- read.delim("spneumoniae_sparc.lineages.tsv")
df.gono <- read.delim("ngonorrhoeae_gisp.lineages.tsv")

all.ants = sort(unique(c(get.ants(df.pneumo), get.ants(df.gono))))
```


# Introduction

In this document, we show how to quantify the association of bacterial clones
with antibiotic resistance. For a given
bacterial species and all antibiotics of interest,
we construct optimal predictors of resistance from lineages and
calculate the associated Receiver Operating Characteristics (ROC)
curve and its Areas under the Curve (AUC).
Comparing the resulting curves and areas among antibiotics helps
to understand the different levels of associations.

We use this framework to show that for the pathogens *Streptococcus pneumoniae* and
*Neisseria gonorrhoeae* antibiotic resistance is highly associated with the population
structure. This provides evidence of the suitability of genomic neighbor
typing as a diagnostic method for these pathogens.


# Optimal lineage-to-resistance classifiers

## Model

Let us have a bacterial species and assume that it has $g$ lineages.
For the purpose of this document, lineages are arbitrary classes of equivalance;
they can be defined based on sequence typing
(e.g., MLST `r citep("10.1146/annurev.micro.59.030804.121325")`)
or clustering (e.g., using BAPS `r citep("10.1093/molbev/mst028")` or PopPUNK `r citep("10.1101/gr.241455.118")`).
Assume that lineages are equally probable; i.e., a randomly drawn isolate $x$
comes from every lineage $i$ with the probability $\frac{1}{g}$.
Assume that for every isolate $x$, we can always correctly determine its lineage $\ell(x)$.
In a clinical setting, this could mean that we can always
determine isolate's MLST sequence type.

Let us consider an antibiotic and assume that every isolate is
either resistant or susceptible to this antibiotic.
Assume that resistance within a lineage $i$ is
iid with the Bernoulli distribution and let $r_i$
denote the probability of resistance. The constants $r_i$
can be determined based on epidemiological data or as proportions
of resistance isolates in individual lineages from population-level
studies.

<!--
Assume that we have a large number of isolates ($N\gg g$) for which we want to
determine resistance.
-->

## Lineage-to-resistance classifiers

For a given species, an antibiotic and fixed probabilities of resistance within
lineages $r_1, \ldots, r_g$,
we construct memoryless probabilistic resistance classifiers $C$ of the form
\[
    C(\ell(x))\to\{\text{'S'},\text{'R'}\}.
\]
In other words, for every isolate we identify its lineage and
the classifier predicts resistance based on the knowledge
of the lineage.

All such classifiers can be parametrized as $C_{(R_1,\ldots,R_g)}$,
where $R_i\in[0, 1]$ is the probability of reporting resistance
given the sample has been indentified as to belong to the lineage $i$.
For instance, the classifier $C_{(0,\ldots,0)}$ always assigns 'S',
$C_{(\frac{1}{2},\ldots,\frac{1}{2})}$ assigns 'R' and 'S' like a fair coin,
$C_{(1,\ldots,1)}$ always assigns 'R', and
$C_{(1,0,\ldots,0)}$ always assigns 'R' for the first lineage and 'S' for the other ones.

With the knowledge of the probabilities of resistance ($r_1, \ldots, r_g$)
within individual lineages $1,\ldots,g$,
we can express false positive rate (FPR)
and true positive rate (TPR)
as a function of the classifier parameters ($R_1, \ldots, R_g$).

```{r chull, fig.width=4, fig.height=4, out.width="40%", fig.cap = "An illustration of the ROC diagram for a species with 4 lineages with resistance probabilities $(r_i)_{i=1}^4=(0.1, 0.4, 0.8, 0.9)$. The grey area corresponds to all the possible lineage-to-resistance classifiers $C_{(R_1,R_2,R_3,R_4)}$ while the red points are the ``vertex\'\' classifiers with integer parameters; i.e., $R_i\\in\\{0,1\\}$ for every $i\\in\\{1,\\ldots,g\\}$. RL denotes the number of lineages for which resistance is reported by an integer classifier, i.e., $\\sum_i R_i$. The optimal classifiers that maximize the AUC lie on the piece-wise linear red curve, which we term the optimal ROC curve."}
source("example-roc-chull.R")
par(mfrow = c(1, 1), mai = c(0.7, 0.8, 0.3, 0.02))
par(ps = 12, cex = 0.8, cex.main = 1)
#par(mgp=c(2,1,0))
#plot.roc.chull(lineages=1, main="1 lineage")
example.roc.chull(lineages = 4,
  p.vec=c(0.1, 0.4, 0.8, 0.9),
  seed = 5,
  main = "ROC curve for a species with 4 lineages")
```


Let us use the standard notation:
\[
  \mathrm{FPR} = \frac{\mathrm{FP}}{\mathrm{FP}+\mathrm{TN}}
  \qquad \text{and} \qquad
  \mathrm{TPR} = \frac{\mathrm{TP}}{\mathrm{TP}+\mathrm{FN}}
\]

where
\begin{align*}
  \mathrm{TP} &= \text{\#true positives; i.e., resistant isolates predicted as resistant} \\
  \mathrm{FP} &= \text{\#false positives; i.e., susceptible isolates predicted as resistant} \\
  \mathrm{FN} &= \text{\#false negatives; i.e., resistant isolates predicted as susceptible} \\
  \mathrm{TN} &= \text{\#true negatives; i.e., susceptible isolates predicted as susceptible}
\end{align*}
Given our assumptions, we can estimate the performance of a classifier $C_{R_1,\ldots,R_g}$ as
\[
\begin{array}{ll}
\mathrm{TP} \approx \frac{N}{g}\sum_{i=1}^g r_iR_i     \qquad &  \mathrm{FP} \approx \frac{N}{g}\sum_{i=1}^g (1-r_i)R_i\\
\mathrm{FN} \approx \frac{N}{g}\sum_{i=1}^g r_i(1-R_i) \qquad &  \mathrm{TN} \approx \frac{N}{g}\sum_{i=1}^g (1-r_i)(1-R_i)
\end{array}
\]
where $N$ is the number of samples tested. We then obtain the following estimates:
\[
\mathrm{FPR} \approx \frac{\sum_{i=1}^g (1-r_i)R_i}{\sum_{i=1}^g (1-r_i)}\\
\qquad \text{and} \qquad
\mathrm{TPR} \approx \frac{\sum_{i=1}^g r_iR_i}{\sum_{i=1}^g r_i}
\]

Using these formulae, we can map every classifier $C_{(R_1,\ldots,R_g)}$
to the ROC space by
\[
  f(R_1, \ldots, R_g) \to \left(
    \frac{\sum_{i=1}^g (1-r_i)R_i}{\sum_{i=1}^g (1-r_i)},
    \frac{\sum_{i=1}^g r_iR_i}{\sum_{i=1}^g r_i}
  \right)
\]

It is easy to see that the map $f$ is linear. Since the set of all possible classifiers is
the cube $[0,1]^g$ in the parameter space, it is convex and its
image in the ROC space must be convex too; an example is provided
in Figure \ref{fig:chull}. Moreover, the image of the cube is equal to the
convex hull of the images of individual cube vertices (red points in Figure \ref{fig:chull}).


## Optimal ROC curves

Our aim now is to find the optimal ROC curve maximizing the AUC (the red curve in Figure \ref{fig:chull}).
Even though the curve corresponds to infinitely
many classifiers, it is a piece-wise linear function which is fully defined
by the $g+1$ ``vertex'' classifiers $C^{(0)}, \ldots, C^{(g)}$ lying on the line intersections.
Enumerating this classifier sequence corresponds to
putting lineages to the order in which they are
switched from susceptible to resistant (i.e., $R_i=0\to R_i=1$) along the red line.
The first classifier $C^{(0)}$, $(0,0)$ in the ROC diagram, corresponds to all
lineages being marked as susceptible, while the last classifier, $C^{(g)}$, $(1,1)$
in the diagram to all lineages being predicted marked as resistant.

The optimal ROC curve can be computed in multiple different ways.
For instance, we can enumerate all cube vertices in the parameter space,
map them to the ROC space, compute the convex hull, and extract its upper part.
More efficiently, we can construct the classifier sequence directly in the ROC
diagram by the following iterative process.
We start at $(0,0)$; i.e., with all lineages marked as susceptible, and at every step
we switch one susceptible lineage to resistant so that
the corresponding step in the ROC graph has maximal possible slope, and
we continue until all lineages are marked as resistant.
If lineages are equally probable, it is easy to see that this order corresponds
to sorting lineages by $r_i$.


```{r pneumo, fig.width=8, fig.height=4, out.width="80%", fig.cap = "\\emph{S. pneumoniae} ROC curves and the corresponding AUCs for ceftriaxone (cro), erythromycin (ery), penicillin (pen), trimethoprim (sxt), and tetracycline (tet)."}
par(mfrow = c(1, 2), mai = c(0.7, 0.8, 0.3, 0.02))
par(ps = 12, cex = 0.8, cex.main = 1)
par(mgp = c(2, 1, 0))
plot.roc(df.pneumo, main="ROC - S. pneumoniae", all.ants = all.ants, norm = 1, ylab = NA)
plot.auc(df.pneumo, main="AUC - S. pneumoniae", all.ants = all.ants, norm = 1, annotate = F)
```


```{r gono, fig.width=8, fig.height=4, out.width="80%", fig.cap = "\\emph{N. gonorrhoeae} ROC curves and the corresponding AUCs for azithromycin (azm), cefixime (cfm), ciprofloxacine (cip), and ceftriaxone (cro)."}
par(mfrow = c(1, 2), mai = c(0.7, 0.8, 0.3, 0.02))
par(ps = 12, cex = 0.8, cex.main = 1)
par(mgp = c(2, 1, 0))
plot.roc(df.gono, main = "ROC - N. gonorrhoeae", all.ants = all.ants, norm = 1, ylab = NA)
plot.auc(df.gono, main = "AUC - N. gonorrhoeae", all.ants = all.ants, norm = 1, annotate = F)
```


# Results

We applied the method to 616 pneumococcal genomes from a carriage study
in Massachusetts children `r citep(c("10.1038/ng.2625","10.1038/sdata.2015.58"))`
and 1102 clinical gonococcal isolates collected from 2000 to 2013
by the Centers for Disease Control and
Prevention’s Gonococcal Isolate Surveillance Project (GISP) `r citep("10.1093/infdis/jiw420")`.
For all isolates, we inferred the resistance categories
as described in the main manuscript, including the
ancestral state responstruction step.
As lineages we used the sequenced clusters
computed using Bayesian Analysis of
Population Structure (BAPS) `r citep("10.1093/molbev/mst028")`.

Lineages of *S. pneumoniae* are predictive for benzylpenicillin,
ceftriaxone, trimethoprim-sulfamethoxazole, erythromycin,
and tetracycline resistance with AUC ranging from 0.90 to 0.97 (Figure \ref{fig:pneumo}). In *N. gonorrhoeae* ciprofloxacin, ceftriaxone, and cefixime attained comparably large AUCs (from 0.93 to 0.98) whereas azithromycin demonstrated lower association (AUC 0.80) (Figure \ref{fig:gono}).


# Bibliography

```{r}
write.bibtex(file="references.bib")
bibliography()
```
